Presheaf Model for Dependent Types

### 2017-09-29 papers/talk-11/paper-11.tex
Update paper11

Now we can define $Ty_F(\Gamma)$ as a set of presheaves over $\int \Gamma$.
Then $A \in Ty_F(\Gamma)$ is a presheaf and for a morphism in $\cat{C}$, $f
\from J \to I$, $\rho \in \Gamma(I)$, $x \in A(I, \rho)$ we have $A(f, \rho)(x)
\in A(J, \Gamma_f \rho)$ since $A$ is a functor. For a context morphism $\sigma
\from \Delta \to \Gamma$ we define a functor $\int \sigma \from \int \Delta \to
\int \Gamma$ by
\[
 \int \sigma (I, \rho') = (I, \sigma_I \rho'),
\]
where $\rho' \in \Delta(I)$.

To define terms of a dependent type $\Gamma \vdash A~type$ we require a family
of elements $a \rho \in A \rho$ for each $I \in \cat{C}$ and $\rho \in
\Gamma(I)$. These elements should satisfy for any $f \from J \to I$ the
equation
\[
  A(f, \rho)(a(I, \rho)) = a(J, \Gamma_f \rho).
\]
The context morphism $\sigma \from \Delta \to \Gamma$ then operates on a term
$\Delta \vdash a\sigma : A\sigma$ by composition
\[
  (a\sigma_I)(I, \rho) = a(I, \sigma_I(\rho)).
\]
We define the presheaf $\Gamma.A$ corresponding to a context extension as
\[
  (\Gamma.A)(I) = \{(\rho, u)~|~\rho \in \Gamma(I) \land u \in A(I, \rho)\}
\]
for objects $I \in \cat{C}$ and for morphisms $f \from J \to I$ as
\[
  (\Gamma.A)_f(\rho, u) = ((\Gamma.A)_f \rho, A(f, \rho)(u)).
\]
The projections are then defined by $\mathbf{p} \from \Gamma.A \to \Gamma$,
$\mathbf{p}(\rho, u) = \rho$, and $\Gamma.A \vdash \mathbf{q} : A\mathbf{p}$ by
$\mathbf{q}(\rho, u) = u$. Under context morphism $\sigma \from \Delta \to
\Gamma$ we define an extension of $\Gamma$ with a term $\Delta \vdash a :
A\sigma$ $(\sigma, a) \from \Delta \to \Gamma.A$ by $(\sigma, a)\rho = (\sigma
\rho, a \rho)$.

@@

So far we know how to represent a type $\Gamma \vdash A~type$ and a dependent
type $\Gamma.A \vdash B$. Now we want to have a representation for a function
that corresponds to a type $\Gamma \vdash \Pi A. B$.

To represent the same type in the model with dependent types we will also use
exponentials, but on the presheaves. Let us consider how to define exponents in
presheaf categories. Let $\Gamma$ and $\Delta$ be presheaves and suppose we
already know how the exponent $\Delta^\Gamma$ is constructed. Then we get by
the Yoneda Lemma and the fact that $-^\Gamma$ should be right adjoint to $-
\times \Gamma$, that
\[
  (\Delta^\Gamma)(I) \cong Hom(\mathbf{y}I, \Delta^\Gamma) \cong Hom(\mathbf{y}I \times \Gamma, \Delta)
\]
where $\mathbf{y}$ is the Yoneda embedding.

>>>>

Now we can define $Ty_F(\Gamma)$ as a set of presheaves over $\int \Gamma$.
Then $A \in Ty_F(\Gamma)$ is a presheaf and for a morphism in $\cat{C}$, $f
\from J \to I$, $\rho \in \Gamma(I)$, $u \in A(I, \rho)$ we have $A(f, \rho)(u)
\in A(J, \Gamma_f \rho)$ since $A$ is a functor. For a context morphism $\sigma
\from \Delta \to \Gamma$ we define a functor $\int \sigma \from \int \Delta \to
\int \Gamma$ by
\[
 \int \sigma (I, \rho') = (I, \sigma_I \rho'),
\]
where $\rho' \in \Delta(I)$.

To define terms of a dependent type $\Gamma \vdash A~type$ we require a family
of elements $a(I, \rho) \in A(I, \rho)$ for each $I \in \ob(\cat{C})$ and $\rho
\in \Gamma(I)$. These elements should satisfy for any $f \from J \to I$ the
equation
\[
  A(f, \rho)(a(I, \rho)) = a(J, \Gamma_f \rho).
\]
The context morphism $\sigma \from \Delta \to \Gamma$ then operates on a term
$\Delta \vdash a\sigma : A\sigma$ by composition
\[
  (a\sigma_I)(I, \rho') = a(I, \sigma_I(\rho')).
\]
for $\rho' \in \Delta(I)$.

We define the presheaf $\Gamma.A$ corresponding to a context extension for
objects $I \in \ob(\cat{C})$ as
\[
  (\Gamma.A)(I) = \{(\rho, u)~|~\rho \in \Gamma(I) \land u \in A(I, \rho)\}
\]
and for morphisms $f \from J \to I$ as
\[
  (\Gamma.A)_f(\rho, u) = ((\Gamma.A)_f \rho, A(f, \rho)(u)).
\]
The projections are then defined by $\mathbf{p} \from \Gamma.A \to \Gamma$,
$\mathbf{p}(\rho, u) = \rho$, and $\Gamma.A \vdash \mathbf{q} : A\mathbf{p}$ by
$\mathbf{q}(\rho, u) = u$. Under context morphism $\sigma \from \Delta \to
\Gamma$ we define an extension of $\Gamma$ with a term $\Delta \vdash a :
A\sigma$, $(\sigma, a) \from \Delta \to \Gamma.A$ by
\[
  (\sigma, a)\rho = (\sigma \rho, a \rho)
\]
for the appropriate subscript on $\sigma$.

The second projection $\mathbf{q}$ here is the one that models the variables in
an extended context. The application of it returns the current term that
substitutes the variable. Note the similarity to the variable rule in the ccc
model of STLC, as it is also modeled by a second projection there.

@@

So far we have a model for types $\Gamma \vdash A~type$ and $\Gamma.A \vdash
B~type$. Now we want to find a model for a dependent function type $\Gamma
\vdash \Pi x:A.B$. This model should connect our representations $A(I, \rho)$
and $B(\rho, u)$ for these types, where $I \in \ob(\cat{C})$, $\Gamma \in
\ob(Psh(\cat{C}))$, $\rho \in \Gamma(I)$, $A \in Psh(\int \Gamma)$, $u \in A(I,
\rho)$.
